% !TEX root = STAG25-MiSo2.tex

\section{Conclusions and future work}
We have introduced TIGHT, an efficient C++ library for correctly rounded interval arithmetic built on top of the state-of-the art in both correctly rounded computations and interval arithmetic.

TIGHT is designed to be future-proof: if correctly rounded mathematical routines become the standard in the coming years, or more performant CR methodologies are developed, switching the underlying library requires minimal changes.
Crucially, since the result of a CR operation is well defined, TIGHT will continue to return the same results \emph{forever} on all machines, even in the event of a library change
%(the only exception to this is, of course bug fixes)
.

The library can still be improved in several ways:
\begin{enumerate}
	\item NFG's arithmetic operations owe their speed to vectorization. Vectorized mathematical libraries exist \cite{sleef} but we are not aware of a correctly-rounded solution. In any case, having non-CR, faster interval routines could still be useful for applications that can cope with error and inconsistencies across machines.
	\item TIGHT currently does not conform with the IEEE 1788-2015 standard for interval arithmetic \cite{ieee1788}. Adding compliance and testing with a framework such as \cite{Benet23} would make the library more easily integrated. This includes adding conformant support for infinity and NaN values.
	\item As mentioned in Section \ref{sub:sincos}, integrating CR functions for multiplication with an irrational constant value as in \cite{Brisebarre:2005} would bring a more consistent speed for $\sin$ and $\cos$ on arbitrary arguments.
	\item CORE-MATH also contains a CR implementation of the gamma function (included in the C++ standard library), that could be used to build an interval implementation, though it is not clear whether this would be useful in practice.
\end{enumerate}

